---
title: "752_21"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## CBB752 Project 2.1 Pipelining

Here we examine the results of two CRISPR off-target prediction tools, CRISPR-Seek (Zhu et al 2014)
and Cas OFFinder (Bae et al 2014). These methods were compared in Haeussler et al (2016). CRISPR Seek 
utilizes an empirically derived substitution matrix to score the likelihood that sites will be cleaved
by the Cas9 nuclease given similarity to the guide (with the substitution matrix quantifying mismatch 
penalties). Cas OFFinder searches the genome for patterns indicative of PAM sites, and scores putative
sites based on base mismatches against the guide query. The following analysis compares the performance 
of these methods against the human reference genome (build 38), and compare against the benchmarking 
results from Haussler et al (2016). 

```{r libraries}
# load libraries
library(tidyr)
library(dplyr)
library(reshape2)
library(CRISPRseek)

library(BSgenome.Hsapiens.UCSC.hg19)
library(TxDb.Hsapiens.UCSC.hg19.knownGene)
library(org.Hs.eg.db)
```

We began by pulling 31 guide RNAs and 650 experimentally validated off target sites from the supplemental data
of Haussler et al (2016). The guides were then be fed into CRISPR-Seek and CasOFFinder to generate
predicted off target sites. These predictions were compared against the validated set to assess each
method's performance.
```{r}
# LOAD IN THE GUIDE RNAS AND VALIDATED OFF TARGET SITES FROM THE PAPER
df <- read.table('13059_2016_1012_MOESM2_ESM.tsv', sep='\t', header=TRUE, stringsAsFactors = FALSE)
# there are 31 unique guide RNAs
# length(unique(df$guideSeq))

# 650 unique off targets
# length(unique(df$otSeq))

# extract guides and off target sequences
guides <- unlist(substr(unique(df$guideSeq),1,20))
offTarget <- unique(df[,c('guideSeq','otSeq')])

# write guide RNAs to file for usage in off target tools
seqNames <- 1:length(guides)
seqNames <- paste0('>', seqNames)
outFile <- paste(seqNames, guides, '', sep='\n')
write(outFile, '~/Desktop/gRNA.fasta')
```

Here we see that CRISPR-Seek predicts 496 sequences which are present in the validated dataset. Checking
for the reverse compliment of CRISPR-Seek prediction increases this number to 516. 
## CRISPR-Seek
```{r CRISPR-Seek}
# GENERATE LIST OF OFF TARGET SITES FROM CRISPRSEEK
# results <- offTargetAnalysis(inputFilePath = '~/Desktop/gRNA.fasta', findgRNAsWithREcutOnly = FALSE, findPairedgRNAOnly = FALSE, findgRNAs = FALSE, BSgenomeName = Hsapiens, txdb = TxDb.Hsapiens.UCSC.hg38.knownGene, orgAnn = org.Hs.egSYMBOL, max.mismatch = 4, outputDir = '~/Desktop/', overwrite = TRUE)

# READ IN RESULTS FROM CRISPRSEEK
CRISPRSEEK <- read.table('OfftargetAnalysis.xls', sep='\t', header=TRUE, stringsAsFactors = FALSE)
list1 <- CRISPRSEEK[,c('OffTargetSequence')]
Rlist1 <- DNAStringSet(list1)
Rlist1 <- Biostrings::reverseComplement(Rlist1)
Rlist1 <- unlist(lapply(Rlist1,toString))

# Check the number of off targets predicted by CRISPR Seek that are in the validated dataset

sum(list1 %in% offTarget$otSeq)
sum(list1 %in% offTarget$otSeq) + sum(Rlist1 %in% offTarget$otSeq)
```

CasOFFinder predicts 461 of the 650 validated off target sites. 
## CasOFFinder
```{r CasOFFinder}
# GENERATE LIST USING CASOFFINDER (www.rgenome.net/cas-offinder)
# this was done using their online tool because R integration is not supported

# READ IN CAS-OFFINDER results
CASOFF <- read.table('result.txt', sep='\t', header=FALSE)
CASOFF_offtargets <- toupper(CASOFF[,3])
RCASOFF_offtargets <- DNAStringSet(CASOFF_offtargets)
RCASOFF_offtargets <- Biostrings::reverseComplement(RCASOFF_offtargets)
RCASOFF_offtargets <- unlist(lapply(RCASOFF_offtargets,toString))

# Check the number of off targets predicted by CAS-OFFinder that are in the validated dataset
length(intersect(CASOFF_offtargets,offTarget$otSeq))
length(intersect(RCASOFF_offtargets,offTarget$otSeq))+ length(intersect(CASOFF_offtargets,offTarget$otSeq))
```

## Comparison against published analyses
The authors of the multi-tool analysis report that CasOFFinder finds all validated off target
sites. Though we ran the tool with the same tolerance of up to 4 mismatches, we did not recover
all validated sites. This may be because we ran our prediction against one version of the genome,
while the curated list of validated off-target sites come from numerous genome versions across
several studies. CRISPR-Seek does slightly better with a total of 516 identified sites, but likely 
suffers from the same limitation (though the authors do not explicitly test the performance
of this tool)
```{r plots, echo=FALSE}
x <- c('CRISPR-Seek','CasOFFinder')
y <- c(516,461)/650
names(y) <- x
barplot(y, ylab='Fraction of Validated Off Target Sites Identified', xlab='Prediction Algorithm', main='Method Comparison')
```


## Citations
1.Zhu, L. J., Holmes, B. R., Aronin, N. & Brodsky, M. H. CRISPRseek: A Bioconductor Package to Identify Target-Specific Guide RNAs for CRISPR-Cas9 Genome-Editing Systems. PLoS One 9, (2014).
2.Bae, S., Park, J. & Kim, J.-S. Cas-OFFinder: a fast and versatile algorithm that searches for potential off-target sites of Cas9 RNA-guided endonucleases. Bioinformatics 30, 1473–1475 (2014).
3.Haeussler, M. et al. Evaluation of off-target and on-target scoring algorithms and integration into the guide RNA selection tool CRISPOR. Genome Biol. 17, 148 (2016).

